evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks

### 2020-08-11 AnalysisOutputLatex.tex
End of day!

%<*ClusterCorrelationSummary0p1>
Moreover, for Stage 2 clusters, spike pair differences were correlated with the spike time of the first unit in the pair (Fisher's method: $\chi^{2}(324, N = 162) = 7832.7, $ $p < .001$; Fig. 3g), showing that correlations were not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
%</ClusterCorrelationSummary0p1>

@@

%<*ClusterCorrelationSummary2>
80.9\% (131/162) of the Stage 2 clusters had correlations of higher significance than correlations calculated for all unclustered first spike pairs in the original response distribution (Fig. 3h). Moreover, 59.9\% (97/162) of the original response distributions from which Stage 2 clusters were extracted were not correlated significantly (p>0.05) (Fig. 3h). 
%</ClusterCorrelationSummary2>

@@

%<*SelectivelyDifferencedTestsPassedNewSummary>
In total, 83 clusters had differencing applied to atleast one unit. 35 of these were correlated and fulfilled the criteria. 23 of these clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig 6b) and less than $45^{\circ}$ (p-value < 0.025; Fig 6b), of which 12 normality was not rejected for (Henze-Zirkler p < 0.05). These angles are similar to the original angles before differencing (Fig. 6c), showing that the differencing operation does not largely alter cluster angle. Fig. 6d shows the amount of variance explained by the pairwise correlation after trend removal. 
%</SelectivelyDifferencedTestsPassedNewSummary>

%<*SelectivelyDifferencedTestsPassedEigenvalueSummary>
The factor analysis model from above can be applied to the 35 correlated criteria fulfilling differenced clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1. 
%</SelectivelyDifferencedTestsPassedEigenvalueSummary>

>>>>

%<*ClusterCorrelationSummary0p1>
spike pair differences were correlated with the spike time of the first unit in the pair for Stage 2 clusters (Fisher's method: $\chi^{2}(324, N = 162) = 7832.7, $ $p < .001$; Fig. 3g), shows that correlations are not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
%</ClusterCorrelationSummary0p1>

@@

%<*ClusterCorrelationSummary2>
80.9\% (131/162) of the Stage 2 clusters had correlations of higher significance than correlations calculated for all unclustered first spike pairs in the originating response distribution (Fig. 3h). Moreover, 59.9\% (97/162) of the original response distributions from which Stage 2 clusters were extracted were not correlated significantly (p>0.05) (Fig. 3h). 
%</ClusterCorrelationSummary2>

@@

%<*SelectivelyDifferencedTestsPassedNewSummary>
In total, 83 clusters had differencing applied to the cluster spike times of at least one unit. 35 of these fulfilled the stationarity and correlation criteria (Methods \ref{methods:angle_and_factor_analysis_criteria}). 23 of these clusters had estimated angles significantly less than $45^{\circ}$ (p-value < 0.025; Fig 6b) and greater than $0^{\circ}$ (p-value < 0.025; Fig 6b), of which for 12 normality was not rejected for (Henze-Zirkler p > 0.05). The differencing operation did not largely alter the cluster angle of the criteria fulfilling differenced clusters (Fig. 6c). Fig. 6d shows the amount of variance explained by the pairwise correlation after trend removal. 
%</SelectivelyDifferencedTestsPassedNewSummary>

%<*SelectivelyDifferencedTestsPassedEigenvalueSummary>
The factor analysis model from above can be applied to the 35 correlated criteria fulfilling differenced clusters. 
%</SelectivelyDifferencedTestsPassedEigenvalueSummary>

### 2020-08-11 manuscript.tex
End of day!

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters observed in first spike response distributions for pairs of units (Fig. X; top left of Supplementary Vid. 1). Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g. flat solid green ellipse in Fig. 3a).
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig. 3).

@@

\ExecuteMetaData[AnalysisOutputLatex.tex]{AngledClustersUniquePairsSummary}
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}

@@

The non-unitary correlations cannot be explained by a model in which the second spiking neuron spikes a fixed time difference after the first. 
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0p1}
This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying process.

@@

This section tests both whether cluster spike times for individual units are affected by neural adaptation and also the degree to which neural adaptation explains cluster correlations.
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitTrialIndexLRSummary}
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitKPSSStationaritySummary}
Only a small amount of variance was explained by the linear models fit to predict the joint spike pair times from trial number (Fig. 4d).

% supporting that neural adaptation does not alone explain cluster correlations. 
% More complex first spike time relationships with trial index were also observed. 
Lag-1 autocorrelations of single unit cluster spike times were above chance [STAT], as expected by the spike times of some units being affected by adaptation.
Beside general adaptive trends, some single units exhibited short trial-by-trial sequences (3-4 trials) of autocorrelated first spike times, before losing such autocorrelations. 
% Lagged autocorrelations were significantly above chance for a high number of lags (Supplementary Fig. 4) supportive of general adaptive trends. 
 Only the lag-2 partial autocorrelation (lags 2-5 tested) was significantly above chance (p-value < 0.005; Supplementary Fig. 4), representative of the minimal and temporary autocorrelations observed for several single units.

Cluster spike times for cluster pairs showed above chance lag-1 cross-correlations, which could be explained by clusters for which the spike times of both units were modulated by adaptation.
Only the lag-2 partial cross-correlation was slightly above chance (lags 2-5 tested) (Supplementary Fig. 5), suggestive of minimal lagged cross-correlative structure between cluster spike times for some unit pairs after neural adaptation.

@@

This section demonstrated heterogeneity in the non-stationarity of Stage 2 clusters. Particularly, either the spike times of both units were not affected by adaptation (i.e. stationary clusters) or one or both of the units have cluster spike times modulated by adaptation (i.e. non-stationary clusters). These two cases must be treated differently when testing for adaptation-independent noise correlations.

@@

Autocorrelations, cross-correlations, partial autocorrelations and partial cross-correlations were at chance levels for the stationary clusters (Supplementary Fig. 7), confirming that the stationary Stage 2 cluster correlations are not caused by neural adaptative trends.

@@

\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} 
% The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters.

@@

The state-conditioned response distributions estimated by factor analysis allow visual (purple ellipses in Fig. 5a and top left of Supplementary Vid. 1) and quantitative (Fig. 5c,d,e; Supplementary Fig. 8) comparison of the estimated state-conditioned response distribution to the larger $4\sigma$ areas under fixed-state decoder assumptions of normality and independence (orange), normal and independent clusters (green flat cluster) and normal and correlated clusters (green angled cluster). The predicted state-conditioned response distributions are strikingly smaller than those predicted under the fixed-state assumptions, thus potentially offering an improved encoding capacity.

@@

[UP TO HERE]

To test whether a shared state could underlie covariation in the remaining non-stationary angled clusters, the non-stationary single unit cluster spike times were differenced and rescaled (a standard time-series analysis technique) to remove neural adaptive trends by subtracting the spike time on trial t by the spike time on trial t-1.
% Clusters which did not fulfil the criteria after differencing were removed from further analysis. 
% Differencing a time series reduces the number of samples by 1 (Methods \ref{methods:angle_and_factor_analysis_criteria}). 
The green angled cluster of Fig. 3a fulfils the criteria after the spike times of its non-stationary unit are differenced (Fig. 6a). Application of the angle estimation technique estimated that the angle was X\todo[color=red!40]{Fill} plus minus Y (Fig. 6a), similar to the original angle and suggestive of an underlying factor after the removal of trial dependence. As expected by differencing, the spike times of the differenced unit show negative lag-1 autocorrelations and negative lag-2 partial autocorrelations (Fig. 6a). 
% Such negative autocorrelations are expected by differencing, and are removed in the next section by fitting ARIMA models.

Due to the expected negative autocorrelations introduced by differencing (Supplementary Fig. 9), the criteria for cluster unit autocorrelation p-values < 0.05, was relaxed for the differenced clusters (criterion 6, Methods \ref{methods:angle_and_factor_analysis_criteria}).
\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedNewSummary}

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedEigenvalueSummary}
Again, the factor analysis angles show close similarity to the PCA angles (Fig. 6e).
The standard deviations of the factor-dependent response distributions (estimated by factor analysis) again offer a strikingly higher potential for encoding than simpler neural coding assumptions (Fig. 6a; Supplementary Vid 1; Fig. 6f,g; Supplementary Fig. 10).

>>>>

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters observed in first spike response distributions for pairs of units (Fig. X; top left of Supplementary Vid. 1). Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g. flat solid green ellipse in Fig. 3a).
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig. 3). The correlated clusters can be analysed to test whether the correlations are explained by adaptative or non-adaptative processes.

@@

\ExecuteMetaData[AnalysisOutputLatex.tex]{AngledClustersUniquePairsSummary}
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}
These results suggest show that clustering is necessary to uncover spike time noise correlations between a large proportion of unit pairs.

@@

Finally, the non-unitary correlation angles and the fact that
% cannot be explained by a model in which the second spiking neuron spikes a fixed time difference after the first. 
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0p1}
This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying process or processes.

@@

This section tests both whether cluster spike times for individual units are affected by neural adaptation. 
% and also the degree to which neural adaptation explains cluster correlations.
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitTrialIndexLRSummary}
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitKPSSStationaritySummary}
Only a small amount of variance was explained by the linear models fit to predict the joint spike pair times from trial number (Fig. 4d).

% supporting that neural adaptation does not alone explain cluster correlations. 
% More complex first spike time relationships with trial index were also observed. 
% Lag-1 autocorrelations of single unit cluster spike times were above chance [STAT], as expected by neural adaptation affecting the spike times of some units.
Beside general adaptive trends, some single units exhibited short and occasional trial-by-trial sequences (3-4 trials) of autocorrelated first spike times.
% Lagged autocorrelations were significantly above chance for a high number of lags (Supplementary Fig. 4) supportive of general adaptive trends. 
Only the lag-2 partial autocorrelation (lags 2-5 tested) was significantly above chance (p-value < 0.005; Supplementary Fig. 4), representative of these short and occasional lagged autocorrelations. Only the lag-2 partial cross-correlation was slightly above chance (lags 2-5 tested) (Supplementary Fig. 5), suggestive of minimal lagged cross-correlative structure between cluster spike times for some unit pairs after neural adaptation. Such highly infrequent and short lag-2 auto and cross-correlations could be ignored when determining cluster stationarity.

% Cluster spike times for cluster pairs showed above chance lag-1 cross-correlations, which could be explained by clusters for which the spike times of both units were modulated by adaptation.

@@

This section demonstrated heterogeneity in the non-stationarity of Stage 2 clusters. Particularly, either the spike times of both units were not modulated by adaptation (i.e. stationary clusters) or the cluster spike times of one or both of the units are adaptation-modulated (i.e. non-stationary clusters). In the latter case, neural adaptation must be accounted for when testing for adaptation-independent noise correlations.

@@

Autocorrelations, cross-correlations, partial autocorrelations and partial cross-correlations were at chance levels for the stationary clusters (Supplementary Fig. 7), confirming that the stationary Stage 2 cluster correlations are not caused by neural adaptative trends.
\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} 
% The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters.

@@

@@

The state-conditioned response distributions estimated by factor analysis allow visual (purple ellipses in Fig. 5a and top left of Supplementary Vid. 1) and quantitative (Fig. 5c,d,e; Supplementary Fig. 8) comparison of the estimated state-conditioned response distribution to the larger $4\sigma$ areas under fixed-state decoder assumptions of normality and independence (orange), normal and independent clusters (green flat cluster) and normal and correlated clusters (green angled cluster). The predicted state-conditioned response distributions are strikingly smaller than those predicted under the fixed-state assumptions, thus offering an improved encoding capacity if downstream neurons can respond in a state-dependent manner.

@@

To test whether a non-adaptative latent state could also modulate the non-stationary Stage 2 clusters, non-stationary single unit cluster spike times were differenced and rescaled (a standard time-series analysis technique) to remove neural adaptive trends (see Methods \ref{methods:differencing}).
% by subtracting the spike time on trial t by the spike time on trial t-1.
% Clusters which did not fulfil the criteria after differencing were removed from further analysis. 
% Differencing a time series reduces the number of samples by 1 (Methods \ref{methods:angle_and_factor_analysis_criteria}). 
As expected by differencing, the spike times of the differenced unit show negative lag-1 and 2 partial autocorrelations (Fig. 6a; Supplementary Fig. 9?). 
% Due to the expected negative autocorrelations introduced as expected by differencing (Supplementary Fig. 9), 
The autocorrelation criterion (criterion 6, Methods \ref{methods:angle_and_factor_analysis_criteria}) for cluster stationarity was therefore relaxed. The following section applies a more complex time series model to remove the negative autocorrelations. The assessment of `raw' differences in this section, however, allows a direct assessment of cluster angles after trend removal.

The green Stage 2 cluster of Fig. 3a fulfils the stationarity and correlation criteria (Methods \ref{methods:angle_and_factor_analysis_criteria}) after the spike times of its non-stationary unit are differenced (Fig. 6a). The $\theta_{45}$ angle was calculated as X\todo[color=red!40]{Fill} plus minus Y (Fig. 6a), similar to the original angle and suggestive of covariation by an underlying state after the removal of the non-stationary unit's adaptative trend.

% Such negative autocorrelations are expected by differencing, and are removed in the next section by fitting ARIMA models.

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedNewSummary}

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedEigenvalueSummary}
Again, the factor analysis angles show close similarity to the PCA angles (Fig. 6e).
The standard deviations of the state-dependent response distributions estimated by factor analysis again offer a strikingly higher potential for encoding than simpler neural coding assumptions (Fig. 6a; Supplementary Vid 1; Fig. 6f,g; Supplementary Fig. 10).
